---
title: "March Madness Mania"
output: pdf_document
date: "2025-03-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Call Libraries
```{r import data, echo=TRUE}
library(tidyverse)
library(dplyr)
```

#Import Dataset
```{r import data, echo=TRUE}
file_path <- file.choose()
ds <- read.csv(file_path)
```

##Import Seed data
```{r import seed data, echo=TRUE}
M_or_W <- "MNCAATourneySeeds.csv"


read_path <- paste0("C:/Users/Evan/Documents/Kaggle/march-machine-learning-mania-2025/", M_or_W)
seed <- read.csv(read_path)

seed <- seed %>%
   mutate(Seed = str_sub(Seed,2) %>% 
            str_replace("[ab]$", ""))

print(seed)
```

##Set Variables
```{r setvariables, echo=TRUE}
szn <- "2025"
```

#Data Manipulation
Want new columns: AVG score margin, FG% differential, eFG% differential, rebound margin, TO margin, 
Important Columns: FT%, ORB%, Opponent TOV%

##Mens 
New: Pace vs Net Rating metric
Important: Kenpom
```{r data manipultion men, echo=TRUE}
seed_filtered <- seed %>%
  filter(Season == szn)

ds <- ds %>%
  mutate(
    Kenpom._.NetRtg = as.numeric(Kenpom._.NetRtg),
    School.Advanced._.Pace = as.numeric(School.Advanced._.Pace)
  )

ds_new <- ds %>%
  left_join(select(seed_filtered, TeamID, Seed), by = c("ID" = "TeamID"))%>%
  mutate(AVG_Score_Margin = (Points._.Tm. - Points._.Opp.) / Overall._.G) %>%
  mutate(FG_pct_Difference = Totals._.FG. - Opponent._.FG.) %>%
  mutate(eFG_pct_Difference = School.Advanced._.eFG. - Opponent.Advanced._.eFG.) %>%
  mutate(Rebound_Margin = School.Advanced._.TRB. - Opponent.Advanced._.TRB.) %>%
  mutate(TO_Margin = School.Advanced._.TOV. - Opponent.Advanced._.TOV.) %>%
  mutate(TriangleEXC = Kenpom._.NetRtg * School.Advanced._.Pace * .01) %>%
  mutate(Season = szn)

```

deal with missing values
```{r missing values, echo=TRUE}
# Remove the ID column from ds before applying PCA
ds_reduced <- ds_new %>%
  select(-ID, -School, -Seed)  # Remove non-numeric or identifier columns

# Check for rows with any NA values
rows_with_na <- ds_reduced %>%
  filter(if_any(everything(), is.na))

# View the rows with NA values
print(rows_with_na)
```

```{r select columns, echo=TRUE}
ds_new <- ds_new %>%
  filter(!is.na(Kenpom._.NetRtg)) %>%
  filter(!is.na(Seed)) %>%
  select(ID, Season, Seed, AVG_Score_Margin, FG_pct_Difference, eFG_pct_Difference, Rebound_Margin, TO_Margin, TriangleEXC, Kenpom._.NetRtg, Kenpom._.SOSNetRtg, Kenpom._.Luck, Totals._.FT., School.Advanced._.ORB., Opponent.Advanced._.TOV., School.Advanced._.eFG., Overall._.W.L.) %>%
  mutate_all(~ as.numeric(as.character(.)))
```

Then reduce demensions with correlation matrix

```{r pca, echo=TRUE}
library(corrplot)

# Calculate the correlation matrix
cor_matrix <- cor(ds_new)

# Visualize the correlation matrix
corrplot(cor_matrix, method = "circle")

```



##Womens
```{r data manipultion women, echo=TRUE}
seed_filtered <- seed %>%
  filter(Season == szn)

ds %>% select(-School) %>%
  mutate_all(~ as.numeric(as.character(.)))

ds_new <- ds %>%
  left_join(select(seed_filtered, TeamID, Seed), by = c("ID" = "TeamID"))%>%
  mutate(AVG_Score_Margin = (Points_Tm. - Points_Opp.) / Overall_G) %>%
  mutate(FG_pct_Difference = Totals_FG. - Opponent_FG.) %>%
  mutate(eFG_pct_Difference = School.Advanced_eFG. - Opponent.Advanced_eFG.) %>%
  mutate(Rebound_Margin = School.Advanced_TRB. - Opponent.Advanced_TRB.) %>%
  mutate(TO_Margin = School.Advanced_TOV. - Opponent.Advanced_TOV.) %>%
  mutate(Season = szn)

```
Missing values
```{r data manipultion women, echo=TRUE}
# Remove the ID column from ds before applying PCA
ds_reduced <- ds_new %>%
  select(-ID, -School, -Seed)  # Remove non-numeric or identifier columns

# Check for rows with any NA values
rows_with_na <- ds_reduced %>%
  filter(if_any(everything(), is.na))

# View the rows with NA values
print(rows_with_na)
```

```{r data manipultion women, echo=TRUE}
ds_new <- ds_new %>%
  filter(!is.na(Seed)) %>%
  select(ID, Season, Seed, AVG_Score_Margin, FG_pct_Difference, eFG_pct_Difference, Rebound_Margin, TO_Margin, School.Advanced_Pace, Overall_W.L., School.Advanced_ORtg, Totals_FT.)
  
```

#Export data
```{r export, echo=TRUE}
save_name <- "M2025c.csv"
save_path <- paste0("C:/Users/Evan/Documents/Kaggle/march-machine-learning-mania-2025/My stats/", save_name)
write.csv(ds_new,save_path,row.names = TRUE)
```




#Merge

```{r load data, echo=TRUE}
#twofour <- read.csv(file.choose())
#compact <- compact %>% filter(Season >= "2011" & Season <= "2024") %>%
#  select(Season,WTeamID,LTeamID) %>% 
#  filter(Season != 2021)
#full_years <- bind_rows(eleven,twelve,thirteen,fourteen,fifteeen,sixteen,seventeen,eighteen,nineteen,twotwo,twothree,twofour)
#full_years <- select(full_years, -X)
```
combine compact and full years
```{r combine, echo=TRUE}
combined <- compact %>% left_join(full_years, by = c("Season", "WTeamID" = "ID")) %>%
  rename_with(~paste0("W_", .), .cols = -c(Season, WTeamID, LTeamID))  # Prefix winning team stats with "W_"

combined <- combined %>%
  left_join(full_years, by = c("Season", "LTeamID" = "ID")) %>%
  rename_with(~paste0("L_", .), .cols = -c(Season, WTeamID, LTeamID, starts_with("W_")))
```
create a win column with winning stats
```{r test, echo=TRUE}
combined_diff <- combined %>%
  mutate(across(starts_with("W_"), 
                ~ . - get(str_replace(cur_column(), "W_", "L_")), 
                .names = "diff_{.col}")) %>%
  select(Season, WTeamID, LTeamID, starts_with("diff_"))  # Keep only differences
combined_diff <- combined_diff %>%
  mutate(Win = 1)  # Since the first team is always the winner in the original data

```

```{r test, echo=TRUE}
tourney_flipped <- combined_diff %>%
  rename_with(~str_replace(., "W_", "temp_"), starts_with("W_")) %>%
  rename_with(~str_replace(., "L_", "W_"), starts_with("L_")) %>%
  rename_with(~str_replace(., "temp_", "L_"), starts_with("temp_")) %>%
  mutate(across(starts_with("W_"), ~ . - get(str_replace(cur_column(), "W_", "L_")), .names = "diff_{.col}")) %>%
  select(Season, WTeamID, LTeamID, starts_with("diff_"))  %>% # Keep only differences 
  mutate(Win = 0)  # Now, the originally losing team is "winning" in this flipped row
```

``` {r target, echo=TRUE}
final_data <- bind_rows(combined_diff, tourney_flipped)
```
drop na values
``` {r na, echo=TRUE}
final_data <- final_data %>%
  drop_na()
```

``` {r export, echo=TRUE}
save_name <- "Mteens.csv"
save_path <- paste0("C:/Users/Evan/Documents/Kaggle/march-machine-learning-mania-2025/My stats/", save_name)
write.csv(teens,save_path,row.names = TRUE)
```

``` {r export, echo=TRUE}
twentys <- final_data %>%
  filter(Season > "2020")
teens <- final_data %>%
  filter(Season < "2020")
```
